### 2015-07-05 narrative/data-description.tex
Scott changes
Lots of edits for clarity.

\subsection{Description of Data}
In order for the middleware and storage layer to understand data,
I/O intentions, and act accordingly, describing data in a way that can
be understood by the system is needed. In this project, this is achieved 
by managing data as \textit{objects}.
An object is the smallest unit of data that consists of raw data and metadata,
and a user space variable, such as \textit{temperature}, may consist of a collection
of objects possibly with different accuracy or resolution.
A key advantage of representing data as \textit{objects}, rather than 
bytes, is that it allows data semantics, user intentions, QoS requirements, and
the relationship between user data to be readily captured and embedded with the 
raw data. As an example, retrieving a 3D field data generated from a simulation and then visualizing it
requires the coordinates and connectivity variable 1) to be accessed together and 2) with low
latency in order to achieve a good user experience. The description of data, 
as enabled by managing data as objects, bridges 
the semantics gap between applications, middleware, and storage systems,
and allows the system to understand user-level data, execute QoS requirements and policies, and
optimize application and system performance.

The description of data is essentially metadata and it may include conventional attributes, such as 
data type, size, dimensionality, relationship to other data, and etc. In particular,
relationship to other data objects can be captured by implementing \textit{soft links}
and different types of relationships can be expressed by different types of soft links. 
In this project, a key metadata attribute
that enables QoS scheduling, data placement, and ???, is \textit{data utility value}, which
captures the priority of objects - it is to our belief that
for exascale science, users must prioritize a small set of objects to be saved on 
higher-level capacity-limited storage layer
to avoid the slow access to large-capacity lower-level storage, such as tapes.
We propose a new technique, described by a utility function plugged in by a user, where objects
can be cast into multiple buckets of data, and each bucket can be prioritized differently according
to its utility value.
Once this is done, the middleware system will construct the description of data, re-organize
and place data to achieve the desired QoS goals and policies.

The user would like the ability to get a certain amount of Quality of Service 
such that they can then make decisions when the expected bandwidth, 
for example is less than what they desire and will then place in a certain set 
of rules which the system can make autonomic decisions to help decide what should be done.
A key insight in this proposed project has been the increased interaction of
the application with the storage system. Towards this end, we propose to
address the quality of service requirements by proivding the application
with mechanisms to specify the quality of I/O service, interrogate the
storage system, and react to the responses. Guided by our past work in
ADIOS \cite{lofstead2008flexible} we propose to explore the design of two
mechanisms for this purpose. Firstly, we will explore the augmentation of
I/O applicatition programing interfaces (I/O APIs) to allow applications to
both specify timing and quality information and also query the storage
system for timing estimates. Through this interface we expect the
application and user to gain insights into how long a single output or input
call will take given the required quality information, and then react to
these estimates by adjusting the quality or restricting the scope of the
data required. Likewise we will explore how an application can provide
timing information to the storage system to allow the storage system to make
optimization decisions to best meet the requirements from the application.

We will also explore external data annotations, such as those
provided by the configuration file in ADIOS. Through the use of these
external augmentation the user can provide insights to the storage system on
the relative value of the data, expected life time and performance
characteristics, as well as relationships between different data sets. With
this information the storage system can make optimizations specific to a use
case. We expect these augmentations to be particularly important for
eviction of data from a storage layer, and migration of data sets to a
different storage layer.

\paragraph{State of the art:} ADIOS implements a binary-packed data format that allows
data characteristics such as min, max and index to be wrapped around data chunks. A
direct benefit is that each data chunk can be operated independently and I/O concurrently
can be maximized. We will build upon this capability and further augment the format
to express data utility. Recently DAMSEL \cite{damsel} provides a rich metadata representation
and management layer that captures the relationship between data blocks for scientific applications.
This allows application data to be mapped to storage system efficiently, without overburdening
users to managing complex data model, such as AMR, from the user space. However, the aspect of 
data importance hasn't been explored, which we believe will be important for exascale storage solutions.

\paragraph{Proposed research approach:} 
We will design and develop new APIs through which users can describe the utility of data
based on their expectations. This can be done by plugging user level functions that will
partition user level data into objects and be executed by the middleware to calculate the 
utility value for each object. The utility value along with other metadata will be encapsulated
by the middleware and utilized by the underlying storage system to management data placement.

We will design an updated I/O API that will expand on the current POSIX I/O
semantics by adding new information to the I/O request. We will explore the
set of application level hints that can be easily provided to the storage
system in order for the storage system to make the most appropriate
optimization decisions. In particular, we expect the application to provide
hints on the length of time before the output data is persistent and
visible to users, and the expected lifetime of the data for data output
calls. For read calls we expect the set of hints to combine latency and
precision requirements. Additional hints will also be investigated.

We will also develop a new type of querying system that allows application
to ask the storage system about completion timing information, the response
being an estimate provided by the storage system. We will explore how these
query functions can be integrated into common applications with minimal code
disruptions, and the set of policies that applications can use to respond to
this new information. We will also explore how applications can query the
storage system to gain insights on available space in different tiers and
how applications can tailor their output data to match the available space. 
Finally we will study the use of an external data annotation system that can
provide information to the storage system without requiring the application
to be recompiled.

\paragraph{Challenges}
The design of new APIs posses both technical and adoption challenges. Here
we will only consider the technical challenges. The proposed set of APIs aim
to expose broad system level environmental information to the application,
and allow the application to adapt dynamically based on this information.
One aspect of this challenge is the design of an interrogative API that
enables a back-and-forth between the application and the storage system to
enable the application to receive enough information to make adaptation
decisions.

>>>>

\subsection{Data Description}
In order for the middleware and storage layer to understand data, I/O
intentions, and act accordingly, data must be described in a way that can be
understood by the system. In this project, we use the smallest chunk of data
stored in the storage system as a \textit{chunk}.  A chunk consists of raw
data and metadata.  A user space variable, such as \textit{temperature}, may
consist of a collection of chunks possibly with different accuracy or
resolution with each chunk representing a different portion of the variable.
A key advantage of having attributes at this lowest level is that data
semantics, user intentions, QoS requirements, and the relationship between user
data to be readily captured and embedded with the raw data so that these
smallest pieces can each be managed independently or collectively, depending on
the need. As an example, retrieving 3D field data generated from a simulation
and then visualizing it requires the coordinates and connectivity variable be
accessed both together and with low latency to achieve a good user experience.
The data description, as enabled by managing data at this smallest level,
bridges the semantics gap between applications, middleware, and storage
systems and allows the system to understand user-level data, execute QoS
requirements and policies, and optimize application and system performance.

%The new APIs will enable the bridge by specifying selectable
%performance/quality/- cost tradeoffs from both the application and system
%perspectives based upon the user guided rules/policy and runtime system
%monitoring status. It allows the middleware to make best possible decisions
%from the feedback of storage system knowledge, such that it will embed user
%intentions and the available system storage.  We want the system to give the
%users a certain amount of currency in terms of bandwidth, storage space on
%each level, and latency expectations. These notions will be fuzzy but they
%will allow the user to make ad-hoc decisions to figure out what needs to be
%saved.

The data description may be stored with the individual chunks or within a
seprate metadata service or both to aid QoS requirements. It would contain
conventional attributes such as data type, size, dimensionality, and the
relationship to other data. In particular, the relationship to other chunks can
be captured by including \textit{typed links} each representing a different
relationship type.  In this project, a key metric stored as a metadata
attribute is \textit{data utility}.  It enables QoS scheduling, data placement
decisions, data lifetime decisions, and captures the chunk priority. It is
our belief that exascale science will require users to prioritize a small set
of chunks to be saved on higher-level capacity-limited storage to avoid the
slow access to large-capacity lower-level storage, such as tape.  We propose a
new technique, described by a utility function provided by a user, where chunks
can be cast into multiple data buckets and each bucket can be prioritized
differently according to its utility value.  Once this is done, the middleware
system will construct the data description and re-organize and place data to
achieve the desired QoS goals and policies.

In the meanwhile, the user would like a certain Quality of Service such that
they can then make decisions about what data to access when.
%for example is less than what they desire and will then place in a certain set
%of rules which the system can make autonomic decisions to help decide what
%should be done.
A key insight in this proposed project has been the increased interaction of
the application with the storage system. Towards this end, we propose to
address the QoS requirements by providing the application with mechanisms to
specify the quality of I/O service, interrogate the storage system, and react
to the responses. Guided by our past work we propose to explore the design of
mechanisms for this purpose.

Our research will address the following questions:
1) How can users interact with the storage system efficiently?,
2) How can users express their intentions into actionable items which can take place in the middleware and/or storage layer?
3) How can users annotate data with a utility metric?

\paragraph{State of the art:}
Our current data annotation techniques demonstrated in
ADIOS~\cite{lofstead:2009:ipdps} implements a binary-packed data format that
allows data characteristics such as min, max and index to be wrapped around
data chunks. A direct benefit is that each data chunk can be operated upon
independently and I/O concurrently can be maximized. We will build upon this
capability and further augment the format to express data utility.  Recently
DAMSEL~\cite{damsel} provides a rich metadata representation and management
layer that captures the relationship between data blocks for scientific
applications.  This allows application data to be mapped to storage system
efficiently, without overburdening users to managing complex data model, such
as AMR, from the user space. However, the aspect of data importance hasn't been
explored, which we believe will be important for exascale storage solutions.
{\color{red}\bf there must be other work in this area, and we need to get more
references}

\paragraph{Proposed research approach:} 
We will design and develop new techniques for data management, description,
and access that allows users to describe the data utility based on their
expectations. This can be done by plugging user level functions that
partition user level data into chunks and be executed by the middleware to
calculate the utility value for each chunk. The utility value along with other
metadata will be encapsulated by the middleware and utilized by the underlying
storage system to management data placement.
%
%We will design an updated I/O API that will expand on the current POSIX I/O
%semantics by adding new information to the I/O request.

We will explore the set of application level hints that can be provided easily
to the storage system. These hints will enable the storage system to make the
most appropriate optimization decisions for this user in the context of the
multi-user environment rather than trying to rely on a one-size-fits-all
approach. In particular, we expect the application to provide hints on the
length of time before the output data is persistent and visible to users and
the expected data lifetime for output sets.  For read calls we expect the set
of hints to combine latency and precision requirements.  Additional hints will
also be investigated.

We will also develop a new type of querying system that allows an application
to ask the storage system about completion timing information. The response
would be an estimated elapsed time for the described operation. We will explore
how these query functions can be integrated into common applications with
minimal code disruptions and the set of policies that applications can use to
respond to this new information. We will also explore how applications can
query the storage system to gain insights on available space in different tiers
and how applications can tailor their output data to match the available space.
Finally we will study the use of an external data annotation system that can
provide information to the storage system without requiring the application to
be recompiled.

%
From an application level, we will explore these techniques through our ADIOS
framework using many of the applications to which we have access, including
XGC1, GTC, and SPECFM3D. We will understand: 1) what hints the users can give
us, 2) how the users can take away much of their application specific code to
``system level code'' through the use of these hints.  We will design an API
that will integrate through the storage layer to give time estimations so that
users can decide which data priorities will be written and read. The utility
value for each chunk will place it into one of ``n'' bins where the first bin
is the most important data and the last bin is the least important data. The
bin will give it a lifetime in the highest level of storage that the user thinks
will be needed. For example: in Figure~\ref{fig:ssio-bucket} we have data that
comes from the XGC1 code.  The user can pick a data-refactoring scheme (see
Sec~\ref{sec:data-refactor}) which will bucket the particle data into 4
prioritization levels.  The user can then pick a data utility, such as 3 months
for the highest priority items and 12 months for the lowest priority data.
The user will then call an API when they specify all of the chunks they want
written (particles, two fields, and a mesh in this example) and the system
will return an estimate of the time it will take to write. The user can then
use another API call which will determine the lowest priority element (in this
case the user will write all of the data (P1-P4). The data will then be stored
on all of the layers of the file system where the two highest priority buckets
will be stored in the Parallel File System, but the next highest priority items
will be stored lower in the storage stack. 
%

\begin{wrapfigure}{r}{0.7\textwidth}
%\vspace{0ex}
        \begin{centering} 
	\includegraphics[scale=0.77]{graphics/SSIO-bucket.pdf}
        %\vspace{1ex}
        \caption{Data is given as a set of chunks which includes utility which contains the lifetime in a storage layer, and the priority of the layer.}
        \label{fig:ssio-bucket}
        \end{centering}
%        \vspace{1ex}
\end{wrapfigure}

We will design new techniques that users can use as part of a data mover, or
the system can automatically use, to raise or lower these sub-chunks up and
down the storage stack. The SSIO layer will then manage the system metadata so
that the user can transparently access any of these sub-chunks without
knowledge of the layer. This requires that when reading the data, the APIs will
then get a time estimate of reading, and then allow the users to determine if
they will read less (only the highest priority sub-chunks) if the time estimate
for reading exceeds their expectations. Existing Sirocco functionality may be
sufficient for these purposes, but it must still be proven in practice.
%

Users will also need to describe information about the data in order to choose
the ``most optimal'' data-refactoring routine. Currently, we understand that
users can tell us that the data represents: 1) phase-space (e.g., the particles
in our example), 2) spatial-changes (e.g., the mesh), 3) space-time (e.g.,
the fields on the mesh). We need this information to get the best possible data
refactoring methods. Although users can request a user-defined data-refactoring
method, system-level methods can use this information to make better decisions.
We will investigate new techniques and new semantics to take this information
and use this to best prioritize and re-organize information.

Finally we will investigate additional information on top of the data model.
This information will be used to indicate data associations that exceed
traditional data models. For example, the fields and mesh in our example have
an inherent relationship. A common data model will describe all of the
variables in the mesh (the coordinates, the units, and the types of elements if
this were a finite-element mesh). We will explore new mechanisms that will
group data together making it explicit that M1, F1, and G1 need to be organized
together.  In otherwords, in will not be helpful if data from M1 and M2 were
organized together along with only F1. When the users wants to visualize data
from the field, F1, they will need only M1 and M2 would be additional
information. Likewise, if the user wants to see F1 and F2, then they will need
the mesh from M1 and M2. This requires that the time to access all of this
data is approximately the same requiring similar access times for all of the
data.  We will investigate what choices the SSIO system must make in order to
keep these pieces ``together'', from a performance perspective. Can the storage
system move everything to a lower layer when all of it doesn't fit or is there
other, less tightly related data can be evicted to preserve the collective
retrieval performance?  We need to investigate the potential problems when this
occurs and how the users can specify additional information to ensure they get
the best possible data from the SSIO layer.

We seek to provide dynamic runtime information to the user and make the
system more transparent. The end user could analyze or visualize data in a more
predictable way since they have confidence how much longer they have to wait
for the data.

%Firstly, we will explore the augmentation of I/O application programing
%interfaces (I/O APIs) to allow applications to both specify timing and quality
%information and also query the storage system for timing estimates. If user
%decide to write or read data to or from hierarchical storage systems, they
%will rely on the API to send their intentions for inquiry and examine the
%system status, including how much data they would like to write/read, desired
%bandwidth, data compression method, etc, or the user can express their
%intension of writing data right now no matter what the traffic is now.
%Through this interface we expect the application and user to gain insights
%into how long a single output or input call will take given the required
%quality information, and then react to these estimates by adjusting the
%quality or restricting the scope of the data required. Likewise we will
%explore how an application can provide timing information to the storage
%system to allow the storage system to make optimization decisions to best meet
%the requirements from the application. 
%
%Secondly, we will also explore external data annotations, such as those
%provided by the configuration file in ADIOS. Through the use of these external
%augmentation the user can provide insights to the storage system on the
%relative value of the data, expected life time and performance
%characteristics, as well as relationships between different data sets. With
%this information the storage system can make optimizations specific to a use
%case. We expect these augmentations to be particularly important for eviction
%of data from a storage layer, and migration of data sets to a different
%storage layer.

\paragraph{Challenges:}
These new techniques offer both technical and adoption challenges. We will only
consider the technical challenges. The proposed techniques and features aim to
expose broad system level environmental information to the application and
allow the application to adapt dynamically based on this information.  One
aspect of this challenge is the design of an interrogative API that offers a
negotiation between the application and the storage system to make adaptation
decisions.  Since we will work closely with many of the leading LCF
applications, we need to make sure that our additional APIs and semantics in
the storage layer will be accepted by these applications and eventually by the
rest of the community. This requires careful examination of all possible
design choices to ensure stable APIs and semantics that do not burden the users
with confusing APIS and semantics that cannot even be used by the system.

### 2015-07-05 narrative/proposed-work.tex
Lots of edits for clarity.

\underline{Principle 1:} A knowledge-centric system , let user knowledge  to define data policies. Today Storage System and I/O
layers are written in a stove-pipe fashion, and quite often don't allow optimizations to take place. We propose to re-design the
layer in a highly integrated fashion where user will place their intentions into the system and actions will dynamically and statically
take place to optimize for both the system and for individual requests.

\underline{Principle 2}: Predictable Performance in the SSIO layer needs to be established so more science can be done on the
exascale Systems. Without predictable performance, the I/O layer performance must be guessed by the users, which quite often
mean that science is not performed well.

\underline{Principle 3:} In a data-centric SSIO system, we need user goals to define the data policies. We introduce data-driven
SSIO optimizations by enabling users to define data manipulations (re-organizations) at a high level and then we express
policy goals by associating intentions (hints) with data;' e.g. ``this data at this level should be accessible in the next three months.''
These high level specifications of desired results will be used in our system for data migration and eviction policies.

>>>>

\underline{Principle 1:} A knowledge-centric system , let user knowledge  to
define data policies. Today Storage System and I/O layers are written in a
stove-pipe fashion, and quite often do not allow optimizations to take place.
We propose to re-design the layer in a highly integrated fashion where user
will place their intentions into the system and actions will dynamically and
statically take place to optimize for both the system and for individual
requests.

\underline{Principle 2}: Predictable Performance in the SSIO layer needs to be
established so more science can be done on the exascale systems. Without
predictable performance, the I/O layer performance must be guessed by the
users, which quite often mean that science is not performed well.

\underline{Principle 3:} In a data-centric SSIO system, we need user goals to
define the data policies. We introduce data-driven SSIO optimizations by
enabling users to define data manipulations (re-organizations) at a high level
and then we express policy goals by associating intentions (hints) with data;
e.g. ``this data at this level should be accessible within 100 seconds in the
next three months.'' These high level specifications of desired results will be
used in our system for data migration and eviction policies.
